Approximation as Differentiation

### 2025-05-08 notes.tex
Enriched hom-functor.
Separate doc for notes.

\documentclass[acmsmall,nonacm]{acmart}

\usepackage{bbm}
\usepackage{tikz-cd}

\geometry{paperwidth=8.3in, paperheight=11.7in} % force to A4 for now
\settopmatter{printacmref=false}

@@

\title{Approximation as Differentiation}
\maketitle

\section{Overview}

Covered so far:
\begin{itemize}
\item $\CMon$-enriched categories (\secref{cmon-enriched})
\item biproducts and semi-additive categories (\secref{biproduct})
\item category $\Fam(I,C)$ of $I$-indexed families of objects of $C$ (\secref{fam})
\item set-indexed products (\secref{set-indexed-product})
\item Grothendieck construction $\Grothendieck{C}F$ and category of families $\Fam(C)$ (\secref{grothendieck})
\item Examples of useful semi-additive categories (\secref{useful-semi-additive-categories}):
   \begin{itemize}
   \item category $\LatGal$ of bounded lattices and Galois connections (\secref{categories-with-biproducts:latgal})
   \item category $\FinVect_F$ of finite-dimensional vector spaces over a field $F$
   (\secref{categories-with-biproducts:fdvect})
   \end{itemize}
\end{itemize}

@@

\input{notes/sec/notation}

\section{Definitions}

\input{notes/sec/cmon-enriched}
\input{notes/sec/biproduct}
\input{notes/sec/fam}
\input{notes/sec/set-indexed-product}
\input{notes/sec/grothendieck}
\input{notes/sec/useful-semi-additive-categories}

% \input{notes/language}

\section{Some references}

\cite{karvonen2020}

\bibliographystyle{plainnat}
\bibliography{bib}

>>>>

\documentclass[acmsmall,screen]{acmart}

\usepackage{tabularx}
\usepackage{bbm}
\usepackage{tikz-cd}

%\geometry{paperwidth=8.3in, paperheight=11.7in} % force to A4 for now
\settopmatter{printacmref=false}
\citestyle{acmauthoryear}
\raggedbottom

@@

\title{Approximation as Differentiation}

\author{Robert Atkey}
\email{robert.atkey@strath.ac.uk}
\orcid{0000-0002-4414-5047}
\affiliation{%
  \institution{University of Strathclyde}
  \city{Glasgow}
  \country{UK}}

\author{Roly Perera}
\email{roly.perera@cl.cam.ac.uk}
\orcid{0000-0001-9249-9862}
\affiliation{%
  \institution{University of Cambridge}
  \city{Cambridge}
  \country{UK}
}
\additionalaffiliation{%
   \institution{University of Bristol}
   \city{Bristol}
   \country{UK}
}

\maketitle

\section{Notes}

\subsection{Overview}

Covered so far:
\begin{itemize}
\item automatic differentiation (\secref{auto-diff})
\item $\CMon$-enriched categories, including enriched functors and enriched Yoneda embedding (\secref{cmon-enriched})
\item biproducts and semi-additive categories (\secref{biproduct})
\item useful semi-additive categories (\secref{useful-semi-additive-categories}):
   \begin{itemize}
   \item category $\LatGal$ of bounded lattices and Galois connections (\secref{categories-with-biproducts:latgal})
   \item category $\FinVect_F$ of finite-dimensional vector spaces over a field $F$
   (\secref{categories-with-biproducts:fdvect})
   \end{itemize}
\item category $\Fam(I,\cat{C})$ of $I$-indexed families of objects of $\cat{C}$ (\secref{fam})
\item set-indexed products (\secref{set-indexed-product})
\item Grothendieck construction $\Grothendieck{\cat{C}}F$ and category of families $\Fam(\cat{C})$
(\secref{grothendieck})
\item useful categories of families for automatic approximation
(\secref{galois-slicing-auto-diff-via-fam}):
   \begin{itemize}
      \item Galois slicing via $\Fam(\LatGal)$ (\secref{galois-slicing-auto-diff-via-fam:galois-slicing})
      \item automatic differentiation via $\Fam(\FinVect)$ (\secref{galois-slicing-auto-diff-via-fam:auto-diff})
   \end{itemize}
\end{itemize}

@@

\input{notes/preliminaries}
\input{notes/auto-diff}
\input{notes/cmon}
\input{notes/cmon-enriched}
\input{notes/biproduct}
\input{notes/useful-semi-additive-categories}
\input{notes/fam}
\input{notes/set-indexed-product}
\input{notes/grothendieck}
\input{notes/auto-diff-galois-slicing-via-fam}
\input{notes/language}

\bibliographystyle{ACM-Reference-Format}
\bibliography{bib}

### 2025-05-08 notes/auto-diff-galois-slicing-via-fam.tex
Separate doc for notes.

$\Fam(\LatGal)$ has:
\begin{itemize}
\item as objects $(X, \partial X)$ all pairs of a set $X$ and and for every $x \in X$, a bounded lattice
$\partial X_x$;
\item as morphisms $(X, \partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of functions $f: X \to
Y$ and families of Galois connections $\partial f_x: \partial X_x \to \partial Y_{f(x)}$ for every $x \in X$.
\end{itemize}

>>>>

$\Fam(\LatGal)$ has:
\begin{itemize}
\item as objects $(X, \partial X)$ all pairs of a set $X$ and and for every $x \in X$, a bounded lattice
$\partial X_x$;
\item as morphisms $(X, \partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of a function $f: X \to
Y$ and for every $x \in X$, a Galois connection $\partial f_x: \partial X_x \to \partial Y_{f(x)}$.
\end{itemize}

### 2025-05-08 notes/auto-diff.tex
Separate doc for notes.

\noindent and again define a corresponding category $\Diff^*$ (equivalent to a full subcategory of
$\Fam(\FinVect^\op)$) of such maps where composition of the cotangent maps respects the backward chain rule.

\todo{Can't combine $\Diff$ and $\Diff^*$ into a single subcategory of $\Fam(\FinVect)$, with morphisms $f: \RR^m \to \RR^n \times (\RR^m \linearto \RR^n) \times (\RR^n \linearto \RR^m)$, because..?}

>>>>

\noindent and again define a corresponding category $\Diff^*$ of such maps where composition of the cotangent
maps respects the backward chain rule. \todo{In practice can't combine $\Diff$ and $\Diff^*$ into a single
category with morphisms $f: \RR^m \to \RR^n \times (\RR^m \linearto \RR^n) \times (\RR^n \linearto \RR^m)$
because..?}

### 2025-05-08 notes/biproduct.tex
Separate doc for notes.

\begin{proposition}
\label{prop:biproduct:from-product-or-coproduct}
Suppose $\cat{C}$ is $\CMon$-enriched.
\begin{itemize}
\item If $\cat{C}$ has biproducts $(X \biprod Y, \biinj_X, \biinj_Y, \biproj_X, \biproj_Y)$ then $(X
\biprod Y, \biproj_X, \biproj_Y)$ is a product and $(X \biprod Y, \biinj_X, \biinj_Y)$ is a coproduct.
\item If $\cat{C}$ has binary products $(X \times Y, \proj_1, \proj_2)$ then:
\begin{enumerate}
\item $\prodM{0_{X,Y}}{0_{X,Z}} = 0_{X,Y \times Z}$
\item $\prodM{f_1 + g_1}{f_2 + g_2} = \prodM{f_1}{f_2} + \prodM{g_1}{g_2}$
\item $(X \times Y, \biinj_X, \biinj_Y, \proj_1, \proj_2)$ is a biproduct with $\biinj_X =
\prodM{\id_X}{0_{X,Y}}: X \to X \times Y$ and $\biinj_Y = \prodM{0_{X,Y}}{\id_Y}: Y \to X \times Y$.
\end{enumerate}
\item If $\cat{C}$ has binary coproducts $(A + B, \inj_1, \inj_2)$ then
\begin{enumerate}
\item $\coprodM{0_{A,C}}{0_{B,C}} = 0_{A + B,C}$
\item $\coprodM{f_1 + g_1}{f_2 + g_2} = \coprodM{f_1}{f_2} + \coprodM{g_1}{g_2}$
\item $(A + B, \inj_1, \inj_2, \biproj_A, \biproj_B)$ is a
biproduct with $\biproj_A: A + B \to A = \coprodM{\id_A}{0_{B,A}}$ and $\biproj_B: A + B \to B =
\coprodM{0_{A,B}}{\id_B}$.
\end{enumerate}
\item If $\cat{C}$ has a terminal (or initial) object then it is a zero object.
\end{itemize}
\end{proposition}

@@

\begin{center}
\begin{tikzcd}[row sep=3em, column sep=3em]
   & A \arrow[d, "\biinj_A", shift left] \arrow[dr, "f_2"] \\
   A' \arrow[r, "\prodM{f_1}{g_1}"] \arrow[dr, "g_1"'] \arrow[ur, "f_1"] & A \biprod B \arrow[u, "\biproj_A", shift left] \arrow[d, "\biproj_B"', shift right] \arrow[r, pos=0.35, "\coprodM{f_2}{g_2}"] & B' \\
   & B \arrow[u, "\biinj_B"', shift right] \arrow[ur, "g_2"']
\end{tikzcd}
\end{center}
\end{proposition}

>>>>

\begin{proposition}
\label{prop:biproduct:from-product-or-coproduct}
Suppose $\cat{C}$ is $\CMon$-enriched.
\begin{itemize}
\item If $\cat{C}$ has biproducts $(X \biprod Y, \biinj_X, \biinj_Y, \biproj_X, \biproj_Y)$ then $(X
\biprod Y, \biproj_X, \biproj_Y)$ is a product and $(X \biprod Y, \biinj_X, \biinj_Y)$ is a coproduct.
\item If $\cat{C}$ has binary products $(X \times Y, \proj_1, \proj_2)$ then:
\begin{enumerate}
\item $\prodM{0_{X,Y}}{0_{X,Z}} = 0_{X,Y \times Z}$
\item $\prodM{f_1 + g_1}{f_2 + g_2} = \prodM{f_1}{f_2} + \prodM{g_1}{g_2}$
\item $(X \times Y, \biinj_X, \biinj_Y, \proj_1, \proj_2)$ is a biproduct with $\biinj_X =
\prodM{\id_X}{0_{X,Y}}: X \to X \times Y$ and $\biinj_Y = \prodM{0_{X,Y}}{\id_Y}: Y \to X \times Y$.
\end{enumerate}
\item If $\cat{C}$ has binary coproducts $(X + Y, \inj_1, \inj_2)$ then
\begin{enumerate}
\item $\coprodM{0_{X,Z}}{0_{Y,Z}} = 0_{X + Y,Z}$
\item $\coprodM{f_1 + g_1}{f_2 + g_2} = \coprodM{f_1}{f_2} + \coprodM{g_1}{g_2}$
\item $(X + Y, \inj_1, \inj_2, \biproj_X, \biproj_Y)$ is a biproduct with $\biproj_X: X + Y \to X =
\coprodM{\id_X}{0_{Y,X}}$ and $\biproj_Y: X + Y \to Y = \coprodM{0_{X,Y}}{\id_Y}$.
\end{enumerate}
\item If $\cat{C}$ has a terminal (or initial) object then it is a zero object.
\end{itemize}
\end{proposition}

@@

\begin{center}
\begin{tikzcd}[row sep=3em, column sep=3em]
   & Y \arrow[d, "\biinj_Y", shift left] \arrow[dr, "f_2"] \\
   X \arrow[r, "\prodM{f_1}{g_1}"] \arrow[dr, "g_1"'] \arrow[ur, "f_1"] & Y \biprod Z \arrow[u, "\biproj_Y", shift left] \arrow[d, "\biproj_Z"', shift right] \arrow[r, pos=0.35, "\coprodM{f_2}{g_2}"] & W \\
   & B \arrow[u, "\biinj_Z"', shift right] \arrow[ur, "g_2"']
\end{tikzcd}
\end{center}
\end{proposition}

### 2025-05-08 notes/language.tex
Separate doc for notes.

We assume a language parameterised by a signature $\Sigma = (S, \Op, \Rel)$ consisting of:
\begin{itemize}
\item a set $S$ of sorts;
\item a set $\Op(s_1, \ldots, s_n; s)$ of operations with input sorts $s_1, \ldots, s_n$ and output sort $s$;
\item a set $\Rel(s_1, \ldots, s_n)$ of $n$-ary relations over sorts $s_1, \ldots, s_n$
\end{itemize}

>>>>

We assume a language parameterised by a signature $\Sigma = (S, \Op, \Rel)$ consisting of:
\begin{itemize}
\item a set $S$ of sorts;
\item a set $\Op(s_1, \ldots, s_n; s)$ of $n$-ary operations with input sorts $s_1, \ldots, s_n$ and output
sort $s$;
\item a set $\Rel(s_1, \ldots, s_n)$ of $n$-ary relations over sorts $s_1, \ldots, s_n$.
\end{itemize}
